---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tb

An experiment to create `data.table-esque` interface integrating all major
tidyverse data transformations, but more compact, and not faster! :).

It is working well now and a fair amount of tests have been written, but expect breaking changes.

## general syntax

The general syntax is `tb1[i, j, ..., by, fill, drop]`. The syntax
`df %tb>% .[i, j, ..., by, fill, drop]` converts `df` to a `tb` transforms it and
returns an object of `df`'s class. I advise to use it, not to add to the confusion of
having to worry which objects in the workspace are standard data frames, tibbles, 
`data.table` objects etc.

* `i` is for row subsetting only, similar to *data.table* but feeding 
a data.frame to it doesn't do a **right join** but a semi join. 
* `j` is for column selection in the SQL sense, so column subsetting with 
optional column creation (similar to `dplyr::select()` and `dplyr::transmute()`).
* `...` is for mutating, summarizing, joining...
* `by` is for mentionning columns to **aggregate** by. Aggregation means we
**consistently** get as many rows as distinct groups, to aggregate without groups, feed a length 0 object to `by`. **NOT** used to "mutate by"!


```{r, message=FALSE}
# setup
library(tb)
library(tidyverse, warn.conflicts = FALSE) # for comparison
```

```{r}
# demo
mtcars %tb>% .[mpg > 31,]
mtcars %tb>% .[disp = mean(disp), drat = mean(drat), by = "cyl"]
```


## Advanced features

* No need to convert a data frame to a `tb`, we can just pipe the input into
`%tb>%` for one call and we'll get back an object of the same class as the input:

```{r}
mtcars %tb>% .[disp = mean(disp), drat = mean(drat), by = "cyl"]
```

* `.` can be used in the `...` so mutating/summarizing arguments can refer to themselves :

```{r}
mtcars %tb>% .[disp = mean(.), drat = mean(.), by = "cyl"]
```

* `?` can be used anywhere (`i`, `j`, `...` or `by`) to select with a 
predicate or a regular expression.

```{r}
iris %tb>% .[Sepal.Length = mean(.), Sepal.Width = mean(.), by = ?is.factor]
```

* `:=` can be used as a replacement for `=` anywhere (if used as a first argument
it is not fed to `i` just like it wouldn't with `=`) but it supports more features,
such as this usage of `.` :

```{r}
mtcars %tb>% .[c("disp", "drat") := mean(.), by = ?is.factor]
```

That makes a convenient `summarize_at` mixed with `summarize`, or `summarize_all`

```{r}
mtcars %tb>% .[(?"^d") := mean(.), wt = median(.), by = ?is.factor]
mtcars %tb>% .[(?"^.*") := max(.), by = c()] # or by = NULL
```

* As long as the lhs is not a symbol it will be evaluated

```{r}
mtcars %tb>% .[paste0("di","sp") := mean(.), by = ?is.factor]
```

* To "mutate by", we use a formula notation, and we prefer to say "along" than
"by" to make the distinction clear with aggregation. We can easily mix it with
ungrouped transformations

```{r}
mtcars %tb>% .[1:4, mean_gear = mean(gear) ~ carb, vs_and_am = vs & am]
```

* We can also "slice along"

```{r}
mtcars %tb>% .[1:2 ~ cyl, ] # with data.table : mtcars_dt[, .SD[1:2], by = cyl]
```

* `:` can be used between col names

```{r}
mtcars %tb>% .[1:2, disp:drat]
```

* `.()` has the same meaning as in `bquote()` not as in data.table !

```{r}
x <- "foo"
y <- c("bar", "baz")
mtcars %tb>% .[1:2, .(x) := .(y)]
```

* `s()` is a selection helper, a bit like `dplyr::vars()`, can be use for
negative subetting or just to spare quotes. 

```{r}
mtcars %tb>% .[1:2,s(-?is.numeric,drat, qsec:am)]
```

* We can access in any argument the current data using`.data`, and the subset of data
with `.subset`, can be useful for advanced filtering :

```{r}
library(matrixStats)
mtcars[1:10, ] %tb>% .[rowAlls(.data > 200),] # rather than `filter_all(mtcars[1:10,], all_vars(. > 200))`
mtcars[1:10, ] %tb>% .[rowAnys(.data > 200),] # rather than `filter_all(mtcars[1:10,], any_vars(. > 200))`
mtcars[1:10, ] %tb>% .[rowAnys(.data[?"^d"] %% 2 == 0),]
# rather than `filter_at(mtcars[1:10], vars(starts_with("d")), any_vars((. %% 2) == 0))`
mtcars[1:10, ] %tb>% .[n = nrow(.subset), by = "cyl"] # counting
```

* unary `+` is used to splice, a bit like `!!!` in the *tidyverse*, except
that it doesn't have to be in `...`, works with any function, and will be 
evaluated in the context of the table. For example to to sort by all :

```{r}
mtcars %tb>% .[1:4,c(1,3,6)][order(+.data),]
mtcars %tb>% .[1:4,c(1,3,6)][order(+-.data),] # descending
mtcars %tb>% .[order(+.data[?"^c"]),][1:4,] # arrange_if
```

* A consequence is that `foo[+bar]` is `cbind(foo, bar)`

```{r}
head(cars, 2) %tb>% .[+head(iris,2)]
```

* We can `rbind()` by using `++` notation in `i` so that `foo[++bar,]` is `rbind(foo, bar)`

```{r}
head(cars, 2) %tb>% .[++tail(cars,2),]
```


* surounding the lhs with `((lhs))` means that the rhs variables should be removed

```{r}
mtcars %tb>% .[1:2,((vs_and_am)) := vs & am]
```

* surrounding the lhs with `{}` means we're not transforming values but transforming names, so renaming and mutating can be mixed.

```{r}
mtcars %tb>% .[1:2,{gear} := "GEAR"]
mtcars %tb>% .[1:2,((GEAR)) := gear] # this is similar
mtcars %tb>% .[1:2,{?"^d"} := toupper(.), DRAT = .*100] #rename_at + mutate
mtcars %tb>% .[1:2,{?"^.*"} := toupper(.)] # rename_all
```

* Spreading is done just by giving a "glue-like" name to our variable :

```{r}
df <- data.frame(x = c("a", "b"), y = c(3, 4), z = c(5, 6))
df
spread(df, x, y)
tb1 <- as_tb(df)
tb1[`{x}` := y, by= "z"]
tb1[`{x}` := y, by= s(-y)] # integrate "x" in by to keep the column
```

We can aggregate as with `reshape` or the `reshape2` package :

```{r}
mtcars %tb>% .['mean_draft_for_vs_{vs}' = mean(drat), by = "am" ]
```

It's very easy to multispread, or even combine with mutating and renaming
statement :

```{r}
mtcars %tb>% .[
  'avg_draft_vs_{vs}' = mean(drat),
  'q2_mpg_cyl_{cyl}' = median(mpg),
  by = "am" ]
```

* we mostly don't need to use functionals such as `lapply`, `mapply`, `map`, `pmap`, 
we can use the `~~` notation instead:

```{r}
# loop on values of vs, am and gear, with a constant value of na.rm
mtcars %tb>% .[1:2, 8:10, foo = mean(c(~~vs, ~~am, ~~gear), na.rm = TRUE)]
# rather than : 
mutate(mtcars[1:2, 8:10], foo = pmap_dbl(list(vs, am, gear), ~mean(c(...), na.rm = TRUE)))

```

* We can use the dollar `$` to pipe, as `tb1$.` returns `tb1` by
convention, it makes tb chains easier to spot, more efficient (less conversions) and
faster to type :

```{r}
mtcars %tb>%
  .[1:20,4:9]$
  .[(?"^.*") := min(.), by= s(vs,am)]
```

