---
title: "`tb` compared to `tidyverse`"
subtitle: "A feature by feature review"
author: "Antoine Fabri, January 2020"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, hygge, ninjutsu]
    nature:
      ratio: 16:10
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, , include = FALSE, cache = FALSE}
knitr::opts_chunk$set(fig.width = 6, message = FALSE, warning = FALSE, comment = "", cache = FALSE, fig.retina = 3, size = 40)
library(flipbookr)
library(tb)

library(matrixStats)
library(tidyverse)
library(withDT)
`%>%` <- function(e1,e2) invisible()
```


# *tb* 

`tb` is an attempt at packing more features into a *data.table* looking interface,
it aims at compact code, flexibility, and takes a different approach to a few concepts.

It takes inspiration from the great data manipulation packages the R community has produced,
but it's also a bit wild, and doesn't really try to respect what you're used to, so
take a big breath!

This deck will thus go through *dplyr* and *tidyr*'s functions and give *tb* equivalents.

It shows *tb*'s output only, the *tidyverse* output being a very close equivalent
or identical.

---

# *tb* objects

A `tb` is a *data.frame* with an additional class `"tb"`, just like a `data.table`
is a *data.frame* with a `"data.table"` class. and a tibble
is a *data.frame* with a `"tbl_df"` class.

    my_tb[i, j, ..., by, fill, drop]

In order not to end up with another type of object in our workspace I use the
`%tb>%` pipe, which allows the *tb* syntax but returns an object of the
original class, so usage becomes.

    my_df %tb>% .[i, j, ..., by, fill, drop]

---

# Peculiarities

* summarizing instructions can be mixed with nesting, packing, and spreading, 
and several spreading instructions can be made in one call.

* renaming instructions can be mixed with mutating instructions (themselves grouped or not)

* `.` is used to refer to the currently mutated or summarized column

* `.()` is used as in `bquote()` (not as in *data.table*!)

* `.data` is used to refer to the data, not `.` like in `magrittr`

* Unary `+` is used for splicing

* `~~` is used before a function argument to signal that the function should
be vectorized on this argument


* `:=` is to be seen as an expansion of `=`, thus in `my_tb[a:=b]`, the first
arg is fed to the dots, not to `i` ! A small exception is that if you want to create a column named `drop` or `by` for instance, you can use `:=` : `my_tb[drop := 42]`




* If we feed only one bracket argument to a tb, it is fed to j as for standard
data frames (list indexing), this is different from what *data.table* does.



---

# dplyr

* `mutate()` family
* `select()` family
* `rename()` family
* `slice()`
* `filter()` family
* `summarize()` family
* `count()` family

---


# mutate() 1/5 : standard use

mutating is done through the `...`

```{r, echo = FALSE}
(data <- data.frame(foo = 1, bar = 2, baz = "a", stringsAsFactors = FALSE ))
```

Create, modify or remove a variable

```{r}
data %>% mutate(foo = foo *2, qux = foo * 10, baz = NULL, corge = list(1:3))
data %tb>% .[foo = .*2, qux = foo * 10, baz = NULL, corge = list(1:3)] #<<
```

*tb* can also create data.frame columns, which `mutate()` can't :

```{r}
data %tb>% .[grault = data.frame(a=1, b=2)] #<<
```

---

# mutate() 2/5 : quasi quotation

We use `.()` as in `bquote()`, which is equivalent with some quasi quotation
features. `.data` can also be used to refers to the current data, it's a *tb* object
though, not a special pronoun object as in the *tidyverse*. 

```{r, echo=FALSE}
data
```

```{r}
foo <- 42
x <- "foo"
y <- quote(foo)
z <- "foo4"
data %>% mutate(foo2 = .data[[x]], foo3 = !!y, !!z := foo, qux = !!foo)
data %tb>% .[foo2 = .data[[x]], foo3 = .(y), .(z) := foo, qux = .(foo)] #<<
```

---

# mutate() 3/5 : scoped variants

```{r, echo = FALSE}
data
```

.pull-left[

`mutate_if()`

```{r}
data %>% mutate_if(is.character, toupper)
data %tb>% .[?is.character := toupper(.)] #<<
```

`mutate_all()`

```{r}
data %>% mutate_all(~paste0("X",.))
data %tb>% .[?".*" := paste0("X",.)] #<<
```

]

.pull-right[

`mutate_at()`

```{r}
data %>% mutate_at(vars(foo,bar), ~.*10)
data %tb>% .[s(foo,bar) := .*10] #<<
```

Putting everything together

```{r}
data %tb>% .[ #<<
  ?is.character := toupper(.), #<<
  s(foo,bar) := .*10,  #<<
  ?".*" := paste0("X",.)] #<<
```

]


---

# mutate() 4/5 : transmute()

`transmute()` is done in `j`, `j` is selection in the *SQL* sense, including
modifications and new columns, similar to what *data.table* proposes. 

.pull-left[

`transmute()`

```{r}
data %>% transmute(bar, qux = foo * 4)
data %tb>% .[s(bar, qux = foo * 4)] #<<
```

`transmute_if()`

```{r}
data %>% 
  transmute_if(is.character, toupper)
data %tb>% 
  .[,?is.character, ?".*" :=toupper(.)] #<<
```


]

.pull-right[

`transmute_all()`

```{r}
data %>% transmute_all(~paste0("X",.))
data %tb>% .[?".*" := paste0("X",.)] #<<
```

`transmute_at`

```{r}
data %>% 
  transmute_at(vars(foo,bar), ~.*10)
data %tb>%
  .[, s(foo,bar), ?".*" := .*10] #<<
```
]

---

# mutate() 5/5 : Window functions

Aggregation and window functions are different things in tb, for window functions use `~`

```{r, echo = FALSE}
(data <- data.frame(foo=c(1,1,2,2), bar = 1:4))
```

.pull-left[
grouped `mutate()`

```{r}
data %>% 
  group_by(foo) %>% 
  mutate(baz = mean(bar))
data %tb>% .[baz = mean(bar) ~ foo] #<<
```
]

.pull-right[
grouped `transmute()`

```{r}
data %>% 
  group_by(foo) %>% 
  transmute(baz = mean(bar))
data %tb>% .[s(baz = mean(bar) ~ foo)] #<<
```
]

---


# select() 1/2


```{r, echo = FALSE}
(data <- data.frame(foo = 1, bar = 2, baz = "a", stringsAsFactors = FALSE ))
```


.pull-left[

Use `s()` as a selection utility.

```{r}
data %>% select(bar, baz)
data %tb>% .[, s(bar, baz)] #<<
```

Use regular expressions to select columns.

```{r}
data %>% select(starts_with("ba"))
data %tb>% .[, ?"^ba"] #<<
```

]

.pull-right[

`s()` can be used to rename too

```{r}
data %>% select(foo, BAR = bar)
data %tb>% .[, s(foo, BAR = bar)] #<<
```

`select_if()` with function or dot expression

```{r}
data %>% select_if(is.numeric)
data %tb>% .[, ?is.numeric] #<<
```
]

---


# select() 2/2


```{r, echo = FALSE}
data
```


.pull-left[

Reorder columns.

```{r}
data %>% select(bar, everything())
data %tb>% .[, s(bar, ?".*")] #<<
```

`select_at()` with renaming

```{r, results}
data %>% select_at(vars(foo),toupper)
data %tb>% .[, s(foo), {?".*"} := toupper(.)] #<<
```
]

.pull-right[

Use `s()` for complex or negative selection.

```{r}
data %>% select(-ends_with("ar"))
data %tb>% .[, s(-?"ar$")] #<<
```

`select_all()` with renaming

```{r}
data %>% select_all(toupper)
data %tb>% .[{?".*"} := toupper(.)] #<<
```

]

---


# rename()



```{r, echo = FALSE}
(data <- data.frame(foo = 1, bar = 2, baz = "a", stringsAsFactors = FALSE ))
```

.pull-left[
`rename()` with `{old} := "new"`

```{r}
data %>% rename(BAR = bar)
data %tb>% .[{bar} := "BAR"] #<<
```

`rename_if()`

```{r}
data %>% 
  rename_if(is.character, toupper)
data %tb>% 
  .[{?is.character} := toupper(.)] #<<
```
]

.pull-right[
`rename_all()`

```{r}
data %>% rename_all(toupper)
data %tb>% .[{?".*"} := toupper(.)] #<<
```

`rename_at()`

```{r}
data %>% 
  rename_at(vars(foo, baz), toupper)
data %tb>% 
  .[{?s(foo, baz)} := toupper(.)] #<<
```
]

---

# slice()


```{r, echo = FALSE}
(data <- data.frame(foo=c(1,1,2,2), bar = 1:4))
```

.pull-left[

```{r}
data %>% slice(c(1, 3, n()))
data %tb>% .[c(1, 3, .N),] #<<
```

]

.pull-right[

```{r}
data %>% 
  group_by(foo) %>% 
  slice(2) %>% 
  ungroup()
data %tb>% .[2 ~ foo,] #<<
```

]

---

# `filter()`

We get help from `library(matrixStats)` here.

```{r, echo = FALSE}
(data <- data.frame(foo=c(1,1,2,2), bar = 1:4, baz = 4:1))
```

.pull-left[

```{r}
data %>% filter(foo == 1)
data %tb>% .[foo ==1,] #<<
```

```{r}
data %>% filter_all(any_vars(. > 3))
data %tb>% .[rowAnys(.data > 3),] #<<
```

]

.pull-right[

```{r}
data %>% filter_at(vars(foo, bar), all_vars(. > 1))
data %tb>% .[rowAlls(.data[s(foo, bar)] > 1),] #<<
```

```{r}
data %>% filter_all(all_vars(. > 1))
data %tb>% .[rowAlls(.data > 1),] #<<
```

]


---

# `filter()`

We get help from `library(matrixStats)` here.

```{r, echo = FALSE}
(data <- data.frame(foo=c(1,1,2,2), bar = 1:4, baz = 4:1))
```

.pull-left[

```{r}
data %>% filter(foo == 1)
data %tb>% .[foo ==1,] #<<
```

```{r}
data %>% filter_all(any_vars(. > 3))
data %tb>% .[rowAnys(.data > 3),] #<<
```

]

.pull-right[

```{r}
data %>% filter_at(vars(foo, bar), all_vars(. > 1))
data %tb>% .[rowAlls(.data[s(foo, bar)] > 1),] #<<
```

```{r}
data %>% filter_all(all_vars(. > 1))
data %tb>% .[rowAlls(.data > 1),] #<<
```

]

---

# bind_rows() and bind_cols()

```{r, echo = FALSE}
(data1 <- data.frame(foo = 1, bar = 2))
(data2 <- data.frame(baz = 3, qux = 4))
```

To `bind_cols()` use unary to splice data frame into mutate argument

```{r}
data1 %>% bind_cols(data2)
data1 %tb>% .[+data2] #<<
```

To `bind_rows()` we use the special `++` notation

```{r}
data1 %>% bind_rows(data1)
data1 %tb>% .[++data1,] #<<
```

---

# arrange()

```{r, echo = FALSE}
(data <- data.frame(foo=c(1,1,2,2), bar = 1:4, baz = 4:1))
```

`arrange()`

```{r}
data %>% arrange(foo, baz)
data %tb>% .[order(foo,baz),] #<<
```

to `arrange_at()` we modify a subset and splice it into `order()`

```{r}
data %>% arrange_at(1:2, sin)
data %tb>% .[order(+lapply(.data[1:2],sin)),] #<<
```

---

# `summarize()` 1/2

```{r, echo = FALSE}
data <- head(mtcars[c("disp", "mpg","vs","am")], 4)
rownames(data) <- NULL
data
```

standard `summarize()`

```{r}
mtcars %>% group_by(vs) %>% summarise(mean = mean(disp), n = n())
mtcars %tb>% .[mean = mean(disp), n = .N, by= vs]
```

To `summarize()` by several colums, use a character vector or a `s()` selection :

```{r}
mtcars %>% group_by(vs, am) %>% summarise(mean = mean(disp), n = n())
mtcars %tb>% .[mean = mean(disp), n = .N, by= s(vs, am)] #<<
```

`summarize()` without grouping, using `by = NULL`

```{r}
mtcars %>% summarise(mean = mean(disp), n = n())
mtcars %tb>% .[mean = mean(disp), n = .N, by= NULL] #<<
```

---

# `summarize()` 2/2 : scoped variants

```{r, echo = FALSE}
(data <- head(as.data.frame(starwars), 4)[c("name","species","height","mass")])
```

To `summarize_at()` we use the same syntax as to `mutate_at()`

```{r}
data %>% 
  group_by(species) %>%
  summarise_at(vars(height, mass), mean)
data %tb>%  .[s(height, mass) := mean(.), by = species] #<<
```


```{r}
data %>% 
  group_by(species) %>%
  summarise_if(is.numeric, mean)
data %tb>% .[?is.numeric := mean(.), by = species] #<<
```

---

# `count()`

```{r, echo = FALSE}
(data <- data.frame(foo=c(1,1,2,2), bar = 1:4, baz = 4:1))
```

.pull-left[

`count()`

```{r}
data %>% count(foo)
data %tb>% .[n = .N, by = "foo"]
```

]

.pull-right[

`add_count()`

```{r}
data %>% add_count(foo)
data %tb>% .[n = .N ~ foo]
```
]

---

# tidyr

* `nest()` / `chop()` / `pack()`
* `spread()`

---

# `nest()`

To `nest()` we summarize into a list

```{r}
(data <- tibble(x = c(1, 1, 1, 2, 2, 3), y = 1:6, z = 6:1))
```

To imitate `nest()` we use `by = NA` to group by unmentionned columns

```{r}
data %>% nest(data = c(y, z))
data %tb>% .[data = tibble(y, z), by = NA]
```

`chop()`

```{r}
data %>% chop(c(y, z))
data %tb>% .[s(y,z) := . , by = NA]
```

To `pack()` we mutate a data frame column, and remove the used variables by
surrounding the `((lhs))` with parentheses. 

```{r}
data %>% pack(foo=c(y,z))
data %tb>% .[((foo)) := data.frame(y,z)]
```


```{r}
as_tibble(iris) %>% nest(data = -Species)
as_tibble(iris) %tb>% .[data = .subset[s(-Species)], by= Species]
```


```{r}
as_tibble(mtcars) %>%
  group_by(cyl) %>%
  nest() %>%
  mutate(models = map(data, ~lm(mpg ~ wt, data = .)))

as_tibble(mtcars) %tb>%
  .[data = .subset[s(-cyl)], by = cyl]$
  .[models = lm(mpg~wt, ~~ data)]
```
