---
title: "Benchmarks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benchmarks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
) 
```

We propose a few benchmarks to compare *tb* to *data.table*, *dplyr* or *tidyr*.
*tb* is not focused on speed, so this is to be seen as a sanity check rather than
a thorough report on *tb*'s speed. 

The conclusion is that *tb* is that *tb* is doing ok and is
generally between *base R* and packages cited above.

We take the *mtcars* dataset and convert it to *tb*, *data.table*, and *tbl_df*, 
then apply our benchmark on this small dataset and on the same, replicated to
be `1e4` times longer.

```{r setup, message = FALSE}
library(knitr)
library(tb)
library(tidyverse)
library(data.table)

mtcars_dt <- as.data.table(mtcars)
mtcars_tb <- as_tb(mtcars)
mtcars_tbl_df <- as_tibble(mtcars)

mtcars2 <- do.call(rbind, replicate(10000, mtcars, F))
mtcars_dt2 <- as.data.table(mtcars2)
mtcars_tb2 <- as_tb(mtcars2)
mtcars_tbl_df2 <- as_tibble(mtcars2)
```

## row subsetting

```{r}
bench::mark(
  time_unit = "us",
  check=FALSE,
  base1 = subset(mtcars, disp < 80),
  base2 = mtcars[mtcars$disp < 80,],
  data.table = mtcars_dt[disp<80],
  tb = mtcars_tb[disp<80,],
  dplyr = filter(mtcars_tbl_df, disp<80)) %>%
  kable()

bench::mark(
  time_unit = "us",
  check=FALSE,
  base1 = subset(mtcars2, disp < 80),
  base2 = mtcars2[mtcars2$disp < 80,],
  data.table = mtcars_dt2[disp<80],
  tb = mtcars_tb2[disp<80,],
  filter(mtcars_tbl_df2, disp<80)) %>%
  kable()
```

## Column subsetting

*tb* is doing well. There is only an overhead for big tables when using 
*data.table* strangely.

```{r}
bench::mark(
  time_unit = "us",
  check=FALSE,
  mtcars[3],
  mtcars_dt[,3],
  mtcars_tb[,3],
  select(mtcars_tbl_df, 3))

bench::mark(
  time_unit = "us",
  check=FALSE,
  mtcars2[3],
  mtcars_dt2[,3],
  mtcars_tb2[,3],
  select(mtcars_tbl_df2, 3))
```

## Creating new columns

```{r}
mtcars_dt_copy <- data.table::copy(mtcars_dt)
mtcars_dt_copy2 <- data.table::copy(mtcars_dt2)
bench::mark(
  time_unit = "us",
  check=FALSE,
  transform(mtcars, mpg_cyl = mpg * cyl),
  mtcars_dt_copy[, mpg_cyl := mpg * cyl],
  mtcars_tb[mpg_cyl = mpg * cyl],
  mutate(mtcars, mpg_cyl = mpg * cyl))

bench::mark(
  time_unit = "us",
  check=FALSE,
  transform(mtcars2, mpg_cyl = mpg * cyl),
  mtcars_dt_copy2[, mpg_cyl := mpg * cyl],
  mtcars_tb2[mpg_cyl = mpg * cyl],
  mutate(mtcars2, mpg_cyl = mpg * cyl))
```

## Aggregating

*tb* is on par with base R, even a bit faster, but not competitive with *data.table* and *dplyr*.

```{r}
bench::mark(
  check=FALSE,
  aggregate(mpg ~ cyl, mtcars, mean),
  mtcars_dt[, .(mpg = mean(mpg)), by = cyl],
  mtcars_tb[,mpg = mean(mpg), .by = "cyl"],
  mtcars_tbl_df %>% group_by(cyl) %>% summarize(mpg= mean(mpg)))

bench::mark(
  check=FALSE,
  aggregate(mpg ~ cyl, mtcars2, mean),
  mtcars_dt2[, .(mpg = mean(mpg)), by = cyl],
  mtcars_tb2[,mpg = mean(mpg), .by = "cyl"],
  mtcars_tbl_df2 %>% group_by(cyl) %>% summarize(mpg= mean(mpg)))
```
