---
title: "tb vs dplyr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tb_vs_dplyr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

In this vignette we go through all the examples of *dplyr*'s data wrangling
functions centered on data frames, and propose an alternative using *tb*.

It's meant as a test of *tb*'s range of possibilities, 
indeed the *tidyverse* has clear functions which aim mostly at doing a single thing well,
so listing its functions is like listing features and is a great test for this
package.

The result of the experiment seem to indicate that *tb* can do most of what *dplyr* can do.

## setup

We'll convert a few well known data sets to *tb*, they are still data frames
so *tidyverse* functions work on them too.

```{r setup}
library(tidyverse, warn.conflicts = F)
library(tb)
library(matrixStats) # for functions `rowAnys()` and `rowAlls()` which will come handy
mtcars <- as_tb(mtcars)
iris   <- as_tb(iris)
starwars <- as_tb(starwars) # from dplyr
```

### `slice()`

|   dplyr                  |            tb           |   data.table         | base                        |
|--------------------------|-------------------------| ---------------------|-----------------------------|
| `slice(mtcars, 1)`       | `mtcars[1,]`            |  `mtcars[1]`         | `head(mtcars, 1)`           |
| `slice(mtcars, n())`     | `mtcars[nrow(.X),]`     |  `mtcars[.N]`        | `tail(mtcars, 1)`           |
| `slice(mtcars, 5:n())`   | `mtcars[5:nrow(.X),]`   |  `mtcars[5:.N]`      | `tail(mtcars, -4)`          |
| `slice(mtcars, -5:-n())` | `mtcars[5:nrow(.X),]`   |  `mtcars[-5:-.N]`    | `mtcars[-5:-nrow(mtcars),]` |
| `slice(mtcars, 1:4)`     | `mtcars[1:4,]`          |  `mtcars[1:4]`       | `mtcars[1:4,]`              |
| `mtcars %>%` <br>&ensp; `group_by(, cyl) %>%` <br>&ensp; `slice(1:2)` | `mtcars[1:2 ~ cyl,]` | `mtcars_dt[, .SD[1:2],`<br>&ensp;`by = cyl]` | `do.call(rbind,lapply(` <br>&ensp; `split(mtcars, mtcars$cyl),`<br>&ensp;`` `[`, 1:2,))`` |

*tb* works a lot like *data.table* or *base R* here, so we added columns for comparison.

A few notable differences :

* We can access the original data with `.X` so we don't need the convenient but "magical" `.N`, we just type `nrow(.X)`.
* We need the extra comma compared to *data.table*, because tb supports list indexing like regular data frames and tibble, so that an unique unnamed argument will be used for column selection.
* We use the `by` argument only for true aggregation, leaving as many rows as distinct combinations of grouping columns. We do however support
manipulations "along" other variables (we try to avoid the word "by"), and this is done by using a 2 sided formula.

## `filter()`

|   dplyr                                                                     |            tb                                            |
|-----------------------------------------------------------------------------|----------------------------------------------------------|
| `filter(starwars, species == "Human")`                                      | `starwars[species == "Human",]`                          |
| `filter(starwars, mass > 1000)`                                             | `starwars[mass > 1000,]`                                 |
| `filter(starwars,` <br>&ensp; `hair_color == "none", eye_color == "black")`              | `starwars[ hair_color == "none" & eye_color == "black",]` |
| `filter(starwars, ` <br>&ensp; `hair_color == "none" | eye_color == "black")`             | `starwars[hair_color == "none" | eye_color == "black",]` |
| `starwars %>% filter(mass > mean(mass, na.rm = TRUE))`                      | `starwars[mass > mean(mass, na.rm = TRUE),]`             |
| `starwars %>%` <br>&ensp; `group_by(gender) %>%` <br>&ensp; `filter(mass > mean(mass, na.rm = TRUE))` | `starwars[mass > mean(mass, na.rm = TRUE) ~ gender,]`    | 

This is again very similar to *data.table*, except for the last example where
we filter along gender, in *data.table* syntax we would write : 

```{r, eval = FALSE}
starwars[, .SD[mass > mean(mass, na.rm = TRUE)], by = gender]
```

```{r}
vars <- c("mass", "height")
cond <- c(80, 150)
```

|   dplyr                                                                     |            tb                                            |
|-----------------------------------------------------------------------------|----------------------------------------------------------|
| `starwars %>%` <br>&ensp; `filter(` <br>&emsp; `.data[[vars[[1]]]] > cond[[1]],` <br>&emsp; `.data[[vars[[2]]]] > cond[[2]])`                                      | `starwars[.X[[vars[[1]]]] > cond[[1]] &` <br>&ensp; `.X[[vars[[2]]]] > cond[[2]],]`                          |

Accessing `.X` is a lot like accessing `.data` with *dplyr*, except that `.X` is really
the table object accessed in the context of the data frame, while `.data` is a special
object that only looks like a data frame when accessed this way.

```{r}
mass <- 80
height <- 150
```


|   dplyr                                              |            tb                                            |
|------------------------------------------------------|----------------------------------------------------------|
| `filter(starwars, mass > !!mass, height > !!height)` | `starwars[mass > .(mass) & height > .(height),]`         |

While `tb` doesn't have a complex unquoting system as `dplyr` does, instead it 
leverages `base::unquote()` and its `.()` notation (which is not to be confused 
with the use that *data.table* makes of it!).

## `filter_all()`

|   dplyr                                 |            tb                |
|-----------------------------------------|------------------------------|
| `filter_all(mtcars, all_vars(. > 150))` | `mtcars[rowAlls(.X > 150),]`  |
| `filter_all(mtcars, any_vars(. > 150))` | `mtcars[rowAnys(.X > 150),]`  |

`rowAll()` and `rowAny()` come from the package `MatrixStats`

We avoid the complexity of the `all_vars()` or `any_vars()` magic by applying
the condition directly to the subset and building the filter from the matrix.

## `filter_at()`

|   dplyr                                 |             tb                                    |
|-----------------------------------------|---------------------------------------------------|
| `filter_at(mtcars, vars(starts_with("d")), any_vars((. %% 2) == 0))` | `mtcars[rowAny(.X[?"^d"] %% 2 == 0),]`  

In this example we keep rows where at least one variable starting with "d" has an even value.

`?` is a special function when evaluated in the context of a `tb`, here it is used with
a string, and the regular expression it contains is applied on the names, returning a logical vector.
We will see it also when looking at `select()`.

This vector is used to subset `.X`, and then like on the example above we compute a boolean matrix
and `rowAny` helps us finish the job.

## `filter_if()`

|   dplyr                                 |             tb                                    |
|-----------------------------------------|---------------------------------------------------|
| `filter_if(mtcars, ~ all(floor(.) == .), all_vars(. != 0))` | `mtcars[rowAll(.X[?all(floor(.) == .)] != 0),]`  

In this example, among "integerish" columns we filter rows with only non zero values.

`?` is this time fed a condition using the `.` placeholder, which is substituted
successively by the elements (columns) of `mtcars`. 

## `arrange()`

|   dplyr                                 |             tb                                    |
|-------------------------------|-----------------------------|
| `arrange(mtcars, cyl, disp)`  | `mtcars[order(cyl,disp),]`  |
| `arrange(mtcars, desc(disp))` | `mtcars[order(-disp),]`     |

To sort character value descending, use `order(-xtfrm(col))`, `xtfrm()` is a base function, or
use `dplyr::desc()` which is exactly the same.

## `arrange_all()`

|   dplyr                                 |             tb                                    |
|-------------------------------|-----------------------------|
| `arrange_all(mtcars)`  | `mtcars[order(+.X),]`        |
| `arrange_all(mtcars, desc)`  | `mtcars[order(+-.X),]` |

Unary `+` means that we splice the input, so `order(+.X)`is equivalent to 
`order(.X[[1]], .X[[2]], <etc>)`.

It is similar to `!!!` in the tidyverse except that :

* The input is spliced once evaluated in the context of the `tb` (so here `.X` is understood!), 
while in the *tidyverse* the input is evaluated in the calling environment.
* It is usable in any function.
* It doesn't necessarily have to be used in the `...` argument.

## `arrange_if()`

|   dplyr                        |            tb                      |
|--------------------------------|------------------------------------|
| `arrange_if(iris, is.numeric)` | `iris[order(+.X[?is.numeric]),]`   |

## `arrange_at()`

```{r}
vars <- c("cyl", "disp")
```


|   dplyr                        |            tb                      |
|--------------------------------|------------------------------------|
| `arrange_at(iris, mtcars)` | `mtcars[order(+.X[vars]),]`   |


## `select()`

|   dplyr                                     |            tb           |
|---------------------------------------------|-------------------------|-|
| `select(iris, starts_with("Petal"))`        | `iris[, ?"^Petal"]`   |
| `select(iris, ends_with("Width"))`          | `iris[, ?"Width$"]`   |
| `select(iris, Species, everything())`       | `iris[, s(Species, ?".*")]` |
| `select(iris, -Sepal.Length, Sepal.Length)` | `iris[, s(-Sepal.Length, Sepal.Length)]` |
| `select(iris, -starts_with("Petal"))` | `iris[, s(-?"^Petal")]` |
| `select(iris, petal_length = Petal.Length)` | `iris[, s(petal_length = Petal.Length)]` | Once transmute is coded!
| `select(iris, obs = starts_with('S'))`| `iris[s({?"^S"} := paste0("obs", seq_along(?"^S")))]` | Once transmute is coded!

We use the `?"<regex>"` notation as a counterpart to the select helpers used
by dplyr.

For complex selection we use `s()`, just like `vars()` in dplyr, `s()` allows
richer syntax. It can be used in many places, to select grouping columns, or
columns to transform, but its main use case is as above.

Another use of `s()` in this same argument is to transform the data, as we'll see in
the `transmute()` section. We already make use of its transforming capabilities in
the last example though.

The syntax of the last example is probably the most surprising as it combines
MANY particularities of the `tb` package. You might want to come back to it later,
but I'll still try to explain it while we're here (and to be fair, the *dplyr*
syntax is far for obvious or well known here as well! I discovered this feature while writing this vignetter): 

* Using `{}` around the left hand side means that we are not assigning to column values,
but to names, and not operating on column values but on names. 
* We have objects of length > 1 on each side, but `:=` allows multi assign in the
style of the `zeallot` package.
* Note that new names are on the right, because it's the side of computed values (unlike in `dplyr::rename()`)

That's a lot to digest but we'll come back to it.


```{r}
set.seed(1)
df <- as.data.frame(matrix(runif(100), nrow = 10))
df <- as_tb(df[c(3, 4, 7, 1, 9, 8, 5, 2, 6, 10)])
```

|   dplyr                               |            tb           |
|---------------------------------------|-------------------------|
| `select(df, V4:V6)`                   | `df[V4:V6]`             |
| `select(df, num_range("V", 4:6))`     | `df[?"V[4-6]"]`         |

```{r}
vars <- list(var1 = sym("cyl"), var2 = sym("am"))
```

|   dplyr                               |            tb           |
|---------------------------------------|-------------------------|
| `select(mtcars, !!!vars)`             | `mtcars[{vars} := names(vars)]` |

```{r}
vars <- c(var1 = "cyl", var2 ="am")
```


## `select_all()`

same as `rename_all` below.

|   dplyr                               |            tb           |
|---------------------------------------|-------------------------|
| `select_all(mtcars, toupper)`              | `mtcars[{?".*"} := toupper(.)]`             |

## `select_at()`

|   dplyr                               |            tb           |
|---------------------------------------|-------------------------|-|
| `select_at(mtcars, vars(-contains("ar"), starts_with("c")), toupper)` | `mtcars[s(-?"ar",?"^c")][{?".*"} := toupper(.)]`

*dplyr* functions are usually designed to do only one thing, the renaming
option of select is an exception, presumably to avoid verbosity. We don't
have the verbosity issue so we do it in two steps.

## `select_if()`

|   dplyr                               |            tb           |
|---------------------------------------|-------------------------|-|
| `select_if(iris, is.numeric)` | `iris[,?is.numeric]`|
| `select_if(iris, is.numeric, toupper)` | `mtcars[?is.numeric][{?".*"} := toupper(.)]`| 

## `rename()`

|   dplyr                               |            tb           |
|---------------------------------------|-------------------------|
| `rename(iris, petal_length = Petal.Length)` | `iris[{"Petal.Length"} := "petal_length"]`|

Another way to rename is `iris[((petal_length)) := Petal.Length]` . The double
parenthesis syntax means that variables seen on the right hand side will be
removed after the operation.

A difference in this case is that the "renamed" column will be at the right end of the data frame.


```{r}
vars <- c(var1 = "cyl", var2 ="am")
```

|   dplyr                               |            tb           |
|---------------------------------------|-------------------------|
| `rename(mtcars, !!vars)` | `mtcars[{vars} := names(vars)]`|


## `rename_all()`

same as `select_all()` above.

|   dplyr                               |            tb           |
|---------------------------------------|-------------------------|
| `rename_all(mtcars, toupper)` | `mtcars[{?".*"} := toupper(.)]` |

## `rename_at()`

|   dplyr                               |            tb           |
|---------------------------------------|-------------------------|
| `rename_at(mtcars, vars(-(1:3)), toupper)` | `mtcars[{-(1:3)} := toupper(.)]` |

## `rename_if()`

|   dplyr                               |            tb           |
|---------------------------------------|-------------------------|
| `rename_if(iris, is.numeric, toupper)` | `iris[{?is.numeric} := toupper(.)]` |

## `mutate()`

|   dplyr                               |            tb           |
|---------------------------------------|-------------------------|
| `mtcars %>%`<br>&ensp;`mutate(``cyl2 = cyl * 2, cyl4 = cyl2 * 2)` | `mtcars[cyl2 = cyl * 2, cyl4 = cyl2 * 2]` |
| `mtcars %>%`<br>&ensp;`mutate(mpg = NULL, disp = disp * 0.0163871)` | `mtcars[mpg = NULL, disp = disp * 0.0163871]` |
| `mtcars %>%`<br>&ensp;`group_by(cyl) %>%`<br>&ensp;`mutate(rank = min_rank(desc(mpg))) %>%`<br>&ensp;`ungroup()` | `mtcars[rank = rank(-mpg, "min") ~ cyl]` | 
| `starwars %>%`<br>&ensp;`group_by(gender) %>%`<br>&ensp;`mutate(mass / mean(mass, na.rm = TRUE)) %>%`<br>&ensp;`pull()` | `starwars[x = mass / mean(mass, na.rm = TRUE) ~ gender][[1]]` 

```{r}
vars <- c("mass", "height")
mass <- 100
```

|   dplyr                               |            tb           |
|---------------------------------------|-------------------------|
| `starwars %>% mutate(`<br>&ensp;`prod = .data[[vars[[1]]]] * .data[[vars[[2]]]])` | `starwars[prod = prod(+.X[vars])]` |
| `mutate(starwars, mass = mass / !!mass)` | `starwars[mass = mass / .(mass)]` |

## `mutate_all()`

|   dplyr                               |            tb           |
|---------------------------------------|-------------------------|
| ` ` | ` ` |

## `mutate_if()`

|   dplyr                               |            tb           |
|---------------------------------------|-------------------------|
| `starwars %>%`<br>&ensp;`mutate_if(is.numeric, scale2, na.rm = TRUE)` | `starwars[(?is.numeric) := scale2(., na.rm = TRUE)]` |
| `iris %>%`<br>&ensp;`mutate_if(is.factor, as.character)` | `starwars[(?is.factor) := as.character(.)]` |
| `iris %>%`<br>&ensp;`mutate_if(is.double, as.integer)` | `starwars[(?is.double) := as.integer(.)]` |

We don't have a simple way to to the following operations at the moment but
once we deal with df columns and unpacking we can do
`iris[(?is.numeric) := tb(scale = scale2(.), log = log(.)), unpack = c(?is.numeric, sep = "_")]` and then unpack with a separator

```{r}
iris %>% mutate_if(is.numeric, lst(max, log))

scale2 <- function(x, na.rm = FALSE) (x - mean(x, na.rm = na.rm)) / sd(x, na.rm)
iris %>% mutate_if(is.numeric, list(~scale2(.), ~log(.)))
iris %>% mutate_if(is.numeric, list(scale = scale2, log = log))
iris %>% mutate_if(is.numeric, list(scale2))
iris %>% mutate_if(is.numeric, list(scale = scale2))
```




## `mutate_at()`

```{r}

```

|   dplyr                               |            tb           |
|---------------------------------------|-------------------------|
| `starwars %>%`<br>&ensp;`mutate_at(c("height", "mass"), scale2)` | `starwars[c("height", "mass") := scale2(.)]` |
| `starwars %>%`<br>&ensp;`mutate_at(c("height", "mass"), scale2, na.rm = TRUE)` | `starwars[c("height", "mass") := scale2(., na.rm = TRUE)]` |
| `starwars %>%`<br>&ensp;`mutate_at(c("height", "mass"), ~scale2(., na.rm = TRUE))` | `starwars[c("height", "mass") := scale2(., na.rm = TRUE)]` |
| `iris %>% mutate_at(vars(matches("Sepal")), log)` | `iris[(?"Sepal") := log(.)]` |


## `transmute()`

|   dplyr                               |            tb           |
|---------------------------------------|-------------------------|
| `mtcars %>%`<br>&ensp;`transmute(displ_l = disp / 61.0237)` | `starwars[c("height", "mass") := scale2(.)]` |


## `transmute_all()`


## `transmute_if()`


## `transmute_at()`

<!-- ## `morph()` -->

<!-- `morph()` was a variant of `mutate()` / `transmute()` considered for *dplyr* -->
<!-- but not implemented so far, which would remove its inputs. -->

## `summarize()`

|   dplyr                               |            tb           |
|---------------------------------------|-------------------------|
| `mtcars %>%`<br>&ensp;`summarise(mean = mean(disp), n = n())` | `mtcars[mean = mean(disp), n = nrow(.X), by = NA]` |
| `mtcars %>%`<br>&ensp;`group_by(cyl) %>%`<br>&ensp;`summarise(mean = mean(disp), n = n())` | `mtcars[mean = mean(disp), n = nrow(.subset), by = "cyl"]` |
| `mtcars %>%`<br>&ensp;`group_by(cyl, vs) %>%`<br>&ensp;`summarise(cyl_n = n())` | `mtcars[cyl_n = nrow(.subset), by = c("cyl", "vs")]` |
| `mtcars %>%`<br>&ensp;`group_by(cyl) %>%`<br>&ensp;`summarise(disp = mean(disp),`<br>&ensp;`sd = sd(disp), double_disp = disp * 2)` | `mtcars[disp = mean(disp), sd = sd(disp),`<br>&ensp;`double_disp = disp * 2, by = "cyl"]` |


```{r}
var <- "mass"
summarise(starwars, avg = mean(.data[[var]], na.rm = TRUE))
starwars[avg = mean(.subset[[var]], na.rm = TRUE), by=c()]
# or
starwars[avg = mean(.(as.symbol(var)), na.rm = TRUE), by=c()]

mass <- 100
summarise(starwars, avg = mean(mass / !!mass, na.rm = TRUE))
starwars[avg = mean(mass / .(mass), na.rm = TRUE), by=NA]

```

## `summarize_all()`

```{r}

# If you want to apply multiple transformations, pass a list of
# functions. When there are multiple functions, they create new
# variables instead of modifying the variables in place:
iris %>%
  group_by(Species) %>%
  summarise_all(list(min, max))

# Note how the new variables include the function name, in order to
# keep things distinct. Passing purrr-style lambdas often creates
# better default names:
iris %>%
  group_by(Species) %>%
  summarise_all(list(~min(.), ~max(.)))

# When that's not good enough, you can also supply the names explicitly:
iris %>%
  group_by(Species) %>%
  summarise_all(list(min = min, max = max))

# When there's only one function in the list, it modifies existing
# variables in place. Give it a name to create new variables instead:
iris %>%
  group_by(Species) %>% 
  summarise_all(list(med = median))

iris %>%
  group_by(Species) %>% 
  summarise_all(list(Q3 = quantile), probs = 0.75)

```

## `summarize_if()`

```{r}
# The _if() variants apply a predicate function (a function that
# returns TRUE or FALSE) to determine the relevant subset of
# columns. Here we apply mean() to the numeric columns:
starwars %>%
  summarise_if(is.numeric, mean, na.rm = TRUE)
```

## `summarize_at()`

```{r}
# The _at() variants directly support strings:
starwars %>%
  summarise_at(c("height", "mass"), mean, na.rm = TRUE)
starwars[c("height", "mass") := mean(.), by = NA]

# You can also supply selection helpers to _at() functions but you have
# to quote them with vars():
starwars %>%
  summarise_at(vars(height:mass), mean, na.rm = TRUE)
starwars[height:mass := mean(.), by = NA]
```

## `count()`, `tally()`, `àdd_count()`, `add_tally()`

```{r}
mtcars %>% tally()
mtcars %>% group_by(cyl) %>% tally()
tb(n=nrow(mtcars)) 

mtcars %>% count(cyl)
mtcars[n = nrow(.subset), by = "cyl"]

mtcars %>% add_tally()
mtcars[n = nrow(.X)]

mtcars %>% add_count(cyl)
mtcars[n = nrow(.subset) ~ cyl]


species <-
 starwars %>%
 count(species, homeworld, sort = TRUE)
species
starwars[n = nrow(.subset), by = c("species", "homeworld")]

species %>% count(species, sort = TRUE)
starwars[n = length(unique(homeworld)), by = "species"]

species <-
 starwars %>%
 count(species, homeworld, sort = TRUE, name = "n_species_by_homeworld")
starwars[n_species_by_homeworld = nrow(.subset), by = c("species", "homeworld")]

species
species %>%
 count(species, sort = TRUE, name = "n_species")
starwars[n_species = length(unique(homeworld)), by = "species"]
# add_count() is useful for groupwise filtering
# e.g.: show details for species that have a single member
starwars %>%
  add_count(species) %>%
  filter(n == 1)
starwars[species %in#% 1, n=1]
starwars[n = nrow(.subset) ~ species][n==1,]
```


> note, we should have L <- NROW, works on list, gives number of rows for data.frames


## `left_join`, `right_join`, `semi_join`, `full_join`, `antijoin`

## `bind_cols` `bind_rows`

## `distinct` `distinct_at` `distinct_all` `distinct_if``



<!-- ## `count()` -->

<!-- ```{r} -->
<!-- # doesn't seem to work right -->
<!-- sw[n = nrow(.), by  = "eye_color"] -->
<!-- ``` -->


<!-- ## `mutate()` -->

<!-- ```{r} -->
<!-- sw[1:4, 1:3, height = height/100] -->
<!-- sw[1:4, 1:3, height_cm = height/100] -->
<!-- sw[1:4, 1:6, height = mean(height) ~ eye_color] -->
<!-- x <- "height_cm" # or x <- quote(height_cm) -->
<!-- sw[1:4, 1:3, {x} := height/100] -->
<!-- sw[1:4, 1:3, `{x}` = height/100] -->
<!-- ``` -->

<!-- ## `mutate_if()` -->

<!-- ```{r, eval = FALSE} -->
<!-- # not supported yet, 2 options, 1st expr below or 2 latter -->
<!-- sw[1:4, 1:3, ?is.numeric := log(..)] -->
<!-- sw[1:4, 1:3, ?is.numeric := log] -->
<!-- sw[1:4, 1:3, ?is.numeric := ~log(.)] -->
<!-- ``` -->

<!-- ## `transmute()` -->

<!-- ```{r} -->
<!-- sw[1:4, height_cm = height/100, .rm = TRUE] -->
<!-- sw[1:4, height = height/100, .rm = TRUE] -->
<!-- ``` -->

<!-- Not completely sure about this one, alternatives would be : -->
<!-- * 1) A special value for `j` : `sw[1:4, NULL, height_cm = height/100]` -->
<!-- * 2) Use `j` as in *data.table* and consider that it is still "subsetting"  -->
<!--   (as `transmute()` is really `SELECT` in SQL): `sw[1:4, slct(height_cm = height/100)]` -->
<!-- * 3) Use `[[` as a `with()` to compute unnamed arguments, so we'd do `sw[[,tb(height_cm = height/100)]]` -->
<!-- * 4) Just use `with()` and we don't need `transmute()` : `with(sw, tb(height_cm = height/100)) -->

<!-- An argument to keep the `.rm` is that it could be `NA` by default so when we use -->
<!-- `by` it acts as if it is `TRUE`, and when mutating acts as if it is `FALSE`. -->

<!-- setting to `FALSE` when summarizing would `nest()` (or rather `chop()`) all remaining columns. -->

<!-- ## `morph()` -->

<!-- `morph()` was a variant of `mutate()` / `transmute()` considered for *dplyr* -->
<!-- but not implemented so far, which would remove its inputs. -->

<!-- ```{r} -->
<!-- sw[1:4, 1:6, (bmi) := mass / ((height/100)^2)] -->
<!-- ``` -->

<!-- ## `summarize()` -->

<!-- ```{r} -->
<!-- sw[mean_height = mean(height, na.rm = TRUE), by = "gender"] -->
<!-- ``` -->

<!-- ## `nest()` -->

<!-- Nesting is a special case of summarizing, but keeping extra columns -->

<!-- ```{r, eval = FALSE} -->
<!-- # not supported yet! -->
<!-- # chopping each column -->
<!-- sw[by=Species, .rm = FALSE] -->
<!-- # nesting relevant columns -->
<!-- sw[Petal = data.frame(Petal.Width, Petal.Length), by="Species"] -->
<!-- ``` -->

<!-- ## `pack()` -->

<!-- Nesting is a special case of summarizing, but keeping extra columns -->

<!-- ```{r, eval = FALSE} -->
<!-- # not supported yet! -->
<!-- sw[Petal = data.frame(Petal.Width, Petal.Length)] -->
<!-- ``` -->

<!-- ## `rename()` -->

<!-- ```{r} -->
<!-- sw[1:4, 1:6, (HEIGHT) := height] -->
<!-- ``` -->

<!-- Not sure if there is room to implement `rename_if()` / `rename_at()`, -->
<!-- they're also not frequent calls so it might be good to be more explicit and use the *dplyr* functions. -->

<!-- ## `spread()` / `pivot_longer` -->

<!-- ```{r, eval = FALSE} -->
<!-- # not supported yet -->
<!-- sw[`mean_height_{eye_color}` = mean(height), by = "gender"] -->
<!-- ``` -->

<!-- ## `gather()` -->

<!-- A call to a function, e.g. `keyval()` will gather the columns in nested data frames. -->

<!-- To end up with a longer table we need then to use the argument `unchop` or `unnest`, which we might rename. -->

<!-- we have several options here too, not necessarily exclusive : -->

<!-- ```{r} -->
<!-- # color is a nested data frame, that we unnest in the same call -->
<!-- # sw[(color) := keyval(hair_color, skin_color, eye_color), .unnest = "color"] -->

<!-- # key and value are packed columns, using zealot semantics, that we unchop in the same call -->
<!-- # sw[c("key", "value") := keyval(hair_color, skin_color, eye_color), .unnest = TRUE] -->
<!-- ``` -->

<!-- ## `bind_rows()`, `` -->

<!-- It's not really an operation "inside of a table" so not sure if it needs to be -->
<!-- supported or how we'd do it. -->

<!-- ## `bind_cols()` -->

<!-- It's equivalent to splicing another table. -->

<!-- Either we make splicing systematic for unnamed args and we can do : -->

<!-- ```{r, eval = FALSE} -->
<!-- tb1[,,tb2] -->
<!-- ``` -->

<!-- Or we splice using unary `+` and do -->

<!-- ```{r, eval = FALSE} -->
<!-- tb1[,, +tb2] -->
<!-- ``` -->

<!-- Another way using or "zeallot like" semantics would be : -->

<!-- ```{r, eval = FALSE} -->
<!-- tb1[names(tb2) := tb2] -->
<!-- ``` -->

<!-- which doesn't require additional functionality but is somewhat redundant. -->

<!-- ## `extract()` -->

<!-- I think we can use `unglue_data()` to replace it completely if we splice : -->

<!-- ```{r, eval = FALSE} -->
<!-- tb[,, +unglue_data(.,col, "{new_col1} foo {new_col2}")] -->
<!-- ``` -->

<!-- ## `left_joins()` -->

<!-- I think only `left_join()` makes sense inside the df. -->

<!-- We could just do (without using the unary `+` to splice): -->

<!-- ```{r, eval = FALSE} -->
<!-- tb1[,, tb2[,cols_tb2]] -->
<!-- ``` -->

<!-- in case of different `by` cols, renaming is not more verbose than `by` specifications : -->

<!-- ```{r, eval = FALSE} -->
<!-- tb1[,, tb2[(new) := old]] -->
<!-- ``` -->

<!-- It's annoying though to have to specify the `by` column if we just want to grab -->
<!-- a couple columns from `tb2`. -->

<!-- So we could define some special helper to work only inside `tb`s. -->

<!-- ```{r, eval = FALSE} -->
<!-- # take 2 columns, no need to mention by cols -->
<!-- tb1[,, lj(tb2, col_a, col_b)] -->
<!-- ``` -->

<!-- Or something like this, where `by_col` is searched in both `tb1` and `tb2` :  -->

<!-- ```{r, eval = FALSE} -->
<!-- tb1[, tb2[,c("col_a","col_b")] ~ by_col] -->
<!-- tb1[, tb2[,c("col_a","col_b")] ~ c(by_col1 = "by_col2")] -->
<!-- ``` -->

<!-- Or maybe a double formula ? -->

<!-- ```{r, eval = FALSE} -->
<!-- # natural left join  -->
<!-- tb1[, ~tb2] -->
<!-- # natural join with column selection -->
<!-- tb1[, tb2 ~ c("col_a","col_b")] -->
<!-- # equivalent to -->
<!-- tb1[, ~tb2[,c("by_col","col_a","col_b")]] -->
<!-- # take all columns, but mention 'by' cols -->
<!-- tb1[, tb2 ~~ by_col] -->
<!-- # mention all -->
<!-- tb1[, tb2 ~ c("col_a","col_b") ~ by_col] -->
<!-- # the most complex case -->
<!-- tb1[, tb2 ~ c("col_a","col_b") ~ c(by_col1 = "by_col2")] -->
<!-- # and we can imagine (but we'll need to have fun parsing bc of precedence) -->
<!-- tb1[, tb2 ~ ?is.numeric ~ c(by_col1 = "by_col2")] -->
<!-- ``` -->

<!-- It might not be that bad because as we have a formula and not a data fame -->
<!-- we can splice automatically all unnamed arguments that are not formulas -->
<!-- (and don't start with `?`). -->

<!-- It's also the more compact we'll get. -->

<!-- ## `complete()`, `expand()`, `fill()` -->

<!-- There could be a `.complete` argument that would be triggered on output but not -->
<!-- much value compared to separate call ? -->

<!-- same for `expand()`, `fill()` -->

<!-- ## `separate()`, `separate_rows()`, `unite()` -->

<!-- `separate()` could be done by using a specific function and the output would -->
<!-- be spliced, but most cases are taken care of by `unglue_data()` very conveniently : -->

<!-- ```{r, eval = FALSE} -->
<!-- tb[, +unglue_data(old, "{this} and {that}")] -->
<!-- ``` -->


<!-- I think `separate_rows()` too, using `unglue_vec(x, "{x}, {x}, {x}", multiple = c)` and unchopping the column in the end. -->

<!-- `unite()` is just `tb[(new) := paste(col1,col2,sep="_")]`  -->

<!-- and we could define `fuse()` as a `paste()` with the ability to remove NAs and  -->
<!-- `_` as a default separator, and `fuse0()` (just for NAs) -->


<!-- ### rolling joins -->

<!-- It's cool to be able to do complicated things in so few keystrokes with -->
<!-- *data.table*, but maybe not so valuable, we could get creative with join notation -->
<!-- but it's not that important, and data.table is already there if needed. -->

<!-- ### mapping -->




<!-- `dplyr::rowwise()` is ugly and imposes to  every mutate call to be rowwise. -->

<!-- `map()` or `lapply()` look bad in mutate calls, though we can get used to them, -->
<!-- and we need `pmap()` or `Map()` when vectorizing on several arguments, and we lose -->
<!-- autocomplete, We can do better. -->

<!-- `Vectorize()` is ugly as well, but is the start to a solution.  -->

<!-- What about marking the argument, so that the function knows we should vectorize -->
<!-- along this one ? The following would nest by row (different alternatives): -->

<!-- ```{r} -->
<!-- # easiest -->
<!-- sw[colors := data.frame(?hair_color, ?skin_color, ?eye_color)]  -->
<!-- # to mark difference with other uses of `?` -->
<!-- sw[colors := data.frame(?+hair_color, ?+skin_color, ?+eye_color)] -->
<!-- # v like vectorize ? -->
<!-- sw[colors := data.frame(v?hair_color, v?skin_color, v?eye_color)] -->
<!-- # or with `:` ? -->
<!-- sw[colors := data.frame(v:hair_color, v:skin_color, v:eye_color)] -->
<!-- sw[colors := data.frame(v:+hair_color, v:+skin_color, v:+eye_color)] -->
<!-- # both following are not too bad! just issue of confusion with splicing and join notation -->
<!-- sw[colors := data.frame(++hair_color, ++skin_color, ++eye_color)] -->
<!-- sw[colors := data.frame(~~hair_color, ~~skin_color, ~~eye_color)]  -->
<!-- sw[colors := data.frame(v~~hair_color, v~~skin_color, v~~eye_color)]  -->
<!-- # this adds brackets, but really does look like something different -->
<!-- sw[colors := data.frame((+hair_color), (+skin_color), (+eye_color))] -->
<!-- # this is more readable but brakets remind of glue -->
<!-- sw[colors := data.frame({+hair_color}, {+skin_color}, {+eye_color})] -->
<!-- # this is readable but probably will be confusing -->
<!-- sw[colors := data.frame(V[hair_color], V[skin_color], V[eye_color])] -->
<!-- ``` -->

<!-- Note that we could define a tag to do `tag$data.frame(?hair_color, ?skin_color, ?eye_color)`, which could be a first step, and nice to use elsewhere. -->

<!-- In `[` we could do without it by looking for `?` everywhere and adapting the parent call -->
<!-- to use `Vectorize()`, but it'll be slow. -->

<!-- A `v:+foo` or `v~~foo` system can be generalized into a broader system of prefixes, -->
<!-- used for joins,for select_if etc... -->

<!-- Opinion now : `~~foo` and `(+foo)` are best. -->
