---
title: "tb vs tidyr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tb_vs_tidyr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

In this vignette we go through all the examples of *tidyr*'s data wrangling
functions centered on data frames, and propose an alternative using *tb*.

It's meant as a test of *tb*'s range of possibilities, 
indeed the *tidyverse* has clear functions which aim mostly at doing a single thing well,
so listing its functions is like listing features and is a great test for this
package.

The result of the experiment seem to indicate that *tb* can do most of what *tidyr* can do.

## setup

We'll convert a few well known data sets to *tb*, they are still data frames
so *tidyverse* functions work on them too.

```{r setup}
library(tidyverse, warn.conflicts = F)
library(tb)
library(unglue)
library(matrixStats) # for functions `rowAnys()` and `rowAlls()` which will come handy
mtcars <- as_tb(mtcars)
iris_tb   <- as_tb(iris)
fish_encounters_tb <- as_tb(fish_encounters)
starwars <- as_tb(starwars) # from dplyr
chickwts <- as_tb(chickwts)
```

## nest

nest_legacy

```{r}
# Nest and unnest are inverses
df <- as_tb(data.frame(x = c(1, 1, 2), y = 3:1))

nest_legacy(df, y)
df[data = tb(y), .by="x"] # or df[data = .subset["y"], .by="x"]

# nesting -------------------------------------------------------------------
nest_legacy(iris, -Species)
iris[data = .subset[s(-Species)], .by = "Species"]


nest_legacy(chickwts, weight)
chickwts[data = tb(weight), .by = "feed"]

```

new nest

```{r}
df  <- tibble(x = c(1, 1, 1, 2, 2, 3), y = 1:6, z = 6:1)
tb1 <- tb(x = c(1, 1, 1, 2, 2, 3), y = 1:6, z = 6:1)
# Note that we get one row of output for each unique combination of
# non-nested variables
df %>% nest(data = c(y, z))
tb1[data = tb(y, z), .by = "x"]

# use tidyselect syntax and helpers, just like in dplyr::select()
df %>% nest(data = one_of("y", "z"))
tb1[data = .subset[c("x", "y")], .by = "x"]

iris %>% nest(data = -Species)
iris_tb[data = .subset[s(-Species)], .by = "Species"]

nest_vars <- names(iris)[1:4]
iris %>% nest(data = one_of(nest_vars))
iris_tb[data = .subset[nest_vars], .by = s(-nest_vars)]

iris %>%
  nest(petal = starts_with("Petal"), sepal = starts_with("Sepal"))
iris[petal = .subset[?"^Petal"], sepal = .subset[?"^Sepal"], .by = "Species"]

iris %>%
  nest(width = contains("Width"), length = contains("Length"))
iris[width = .subset[?"Width"], length = .subset[?"Length"], .by = "Species"]
# Nesting a grouped data frame nests all variables apart from the group vars
library(dplyr)
fish_encounters %>%
  group_by(fish) %>%
  nest()
fish_encounters[data = .subset[?".*"], .by = "fish"]
```


## unnest

unnest_legacy

```{r}

# unnesting -----------------------------------------------------------------
df <- tibble(
  x = 1:2,
  y = list(
    tibble(z = 1),
    tibble(z = 3:4)
  )
)
df %>% unnest_legacy(y)

# You can also unnest multiple columns simultaneously
df <- tibble(
  a = list(c("a", "b"), "c"),
  b = list(1:2, 3),
  c = c(11, 22)
)
df %>% unnest_legacy(a, b)
# If you omit the column names, it'll unnest all list-cols
df %>% unnest_legacy()

# unnest() is primarily designed to work with lists of data frames
df <- tibble(
  x = 1:3,
  y = list(
    NULL,
    tibble(a = 1, b = 2),
    tibble(a = 1:3, b = 3:1)
  )
)
df %>% unnest(y)
df %>% unnest(y, keep_empty = TRUE)

# If you have lists of lists, or lists of atomic vectors, instead
# see hoist(), unnest_wider(), and unnest_longer()

#' # You can unnest multiple columns simultaneously
df <- tibble(
 a = list(c("a", "b"), "c"),
 b = list(1:2, 3),
 c = c(11, 22)
)
df %>% unnest(c(a, b))

# Compare with unnesting one column at a time, which generates
# the Cartesian product
df %>% unnest(a) %>% unnest(b)
```

## chop

```{r}
df  <- tibble(x = c(1, 1, 1, 2, 2, 3), y = 1:6, z = 6:1)
tb1 <- tb(x = c(1, 1, 1, 2, 2, 3), y = 1:6, z = 6:1)
df %>% chop(c(y, z))
tb1[y = list(y), z = list(z), .by = "x"]
tb1[s(y, z) := list(.), .by = "x"]
```

## unchop

```{r}
# Unchop ============================================================
df <- tibble(x = 1:4, y = list(integer(), 1L, 1:2, 1:3))
df %>% unchop(y)
df %>% unchop(y, keep_empty = TRUE)

# Incompatible types -------------------------------------------------
# If the list-col contains types that can not be natively
df <- tibble(x = 1:2, y = list("1", 1:3))
try(df %>% unchop(y))
df %>% unchop(y, ptype = tibble(y = integer()))
df %>% unchop(y, ptype = tibble(y = character()))
df %>% unchop(y, ptype = tibble(y = list()))

# Unchopping data frames -----------------------------------------------------
# Unchopping a list-col of data frames must generate a df-col because
# unchop leaves the column names unchanged
df <- tibble(x = 1:3, y = list(NULL, tibble(x = 1), tibble(y = 1:2)))
df %>% unchop(y)
df %>% unchop(y, keep_empty = TRUE)
```

## pack

```{r}
# Packing =============================================================
# It's not currently clear why you would ever want to pack columns
# since few functions work with this sort of data.
df <- tibble(x1 = 1:3, x2 = 4:6, x3 = 7:9, y = 1:3)
tb1 <- tb(x1 = 1:3, x2 = 4:6, x3 = 7:9, y = 1:3)
df
df %>% pack(x = starts_with("x"))
tb1[x = list(.X[?"^x"])][s(-?"x\\d")]
df %>% pack(x = c(x1, x2, x3), y = y)
```


## unpack

```{r}
# Unpacking ===========================================================
df <- tibble(
  x = 1:3,
  y = tibble(a = 1:3, b = 3:1),
  z = tibble(X = c("a", "b", "c"), Y = runif(3), Z = c(TRUE, FALSE, NA))
)
df
df %>% unpack(y)
df %>% unpack(c(y, z))
df %>% unpack(c(y, z), names_sep = "_")

```

## extract

```{r}
df <- data.frame(x = c(NA, "a-b", "a-d", "b-c", "d-e"))
tb <- as_tb(df)
df %>% extract(x, "A")
tb[,,((A)) := unglue_vec(x, "{A}-{B}")]
tb[,,+unglue_data(x, "{A}-{=.*?}")]

df %>% extract(x, c("A", "B"), "([[:alnum:]]+)-([[:alnum:]]+)")
tb[,,+unglue_data(x, "{A}-{B}")]
tb[((c("A","B"))) := unglue_data(x, "{A}-{B}")]

# If no match, NA:
df %>% extract(x, c("A", "B"), "([a-d]+)-([a-d]+)")
tb[((c("A","B"))) := unglue_data(x, "{A=[a-d]+}-{B=[a-d]+}")]
```

## separate

```{r}
df <- data.frame(x = c(NA, "a.b", "a.d", "b.c"))
tb <- as_tb(df)
df %>% separate(x, c("A", "B"))
tb[,,+splt(x,c("A","B"))]
tb[((c("A","B"))) := splt(x)]
# If you just want the second variable:
df %>% separate(x, c(NA, "B"))

# If every row doesn't split into the same number of pieces, use
# the extra and fill arguments to control what happens
df <- data.frame(x = c("a", "a b", "a b c", NA))
df %>% separate(x, c("a", "b"))
tb <- as_tb(df)
tb[,,+splt(x,c("a","b"))][s(-x)]
tb[((c("a","b"))) := splt(x)]

# The same behaviour drops the c but no warnings
df %>% separate(x, c("a", "b"), extra = "drop", fill = "right")
tb[,, +splt(x, c("a", "b"), extra = "drop", fill = "right")]
# Another option:
df %>% separate(x, c("a", "b"), extra = "merge", fill = "left")
tb[,, +splt(x, c("a", "b"), extra = "merge", fill = "left")]
# Or you can keep all three
df %>% separate(x, c("a", "b", "c"))
tb[,, +splt(x, c("a", "b", "c"))]

# If only want to split specified number of times use extra = "merge"
df <- data.frame(x = c("x: 123", "y: error: 7"))
df %>% separate(x, c("key", "value"), ": ", extra = "merge")
tb <- as_tb(df)
tb[,,+splt(x,c("key","value"), extra = "merge")][s(-x)]

# Use regular expressions to separate on multiple characters:
df <- data.frame(x = c(NA, "a?b", "a.d", "b:c"))
df %>% separate(x, c("A","B"), sep = "([\\.\\?\\:])")
tb <- as_tb(df)
tb[c("A","B") := splt(x, sep = "([\\.\\?\\:])")][s(-x)]

# convert = TRUE detects column classes
df <- data.frame(x = c("a:1", "a:2", "c:4", "d", NA))
df %>% separate(x, c("key","value"), ":") %>% str
df %>% separate(x, c("key","value"), ":", convert = TRUE) %>% str
tb <- as_tb(df)
tb[,,+splt(x,c("key","value"), convert = TRUE)][s(-x)]

# Argument col can take quasiquotation to work with strings
var <- "x"
df %>% separate(!!var, c("key","value"), ":")
tb[,,+splt(.(as.symbol(var)),c("key","value"), convert = TRUE)][s(-.(as.symbol(var)))]
```

## separate_rows

```{r}
df <- data.frame(
  x = 1:3,
  y = c("a", "d,e,f", "g,h"),
  z = c("1", "2,3,4", "5,6"),
  stringsAsFactors = FALSE
)
tb <- as_tb(df)
separate_rows(df, y, z, convert = TRUE)
tb[s(y,z) := strsplit(., ",")] # needs to be exanded in later step
```

